Evaluating FAIR Digital Object as a distributed object system

### 2023-01-25 latex/main.tex
more regex!

The FAIR Digital Object Forum (\cite{DMEBhHID}'}}) working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}. As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public. As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:

@@

\textbf{Upload of FDO} \cite{1H7cmmr69} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve). ResourceSync (\cite{eS5y9TRh}}) is suggested as one type of service to list FDOs. This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).

@@

The concept of \textbf{FAIR Digital Objects} \cite{IHLT6hye} has been introduced as way to expose research data as active objects that conform to the FAIR principles \cite{6DjakjNS}. This builds on the \emph{Digital Object} (DO) concept \cite{11MnuwJ4l}, first introduced in 1995 \cite{3Uqe3fuK} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles. DO was the inspiration for the ITU X.1255 framework (\cite{103Hw8H43}}) which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 (\cite{16uB3jxpa} DONA Foundation'}}).

@@

\begin{itemize}
\tightlist
\item
  Building on the Digital Object concept, using the simplified DOIP v2 specification \cite{13TcbsZF6}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 \cite{13TcbsZF6}, see (\cite{xuQKRRx5}}).} (usually TCP/IP over TLS). The main DOIP operations are retrieving, creating and updating digital objects. These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}. FDO types are registered in the local Cordra instance, where they are specified using JSON Schema (\cite{15EZ2D0Rm}}) and PIDs are assigned using the Handle system. Several type registries have been established.
\item
  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
\item
  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
\end{itemize}

@@

Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) (\cite{16WGmCih8}}) that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format (\cite{16wvFH4nh}}) being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).

While triple-based knowledge representations were not new \cite{NxvhwSmG}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs \cite{RRtwPF7Y} are generalised forms of URLs that include locator-less identifiers
  such as ISBN book numbers (URNs). The distinction between locator-full and locator-less identifiers have weakened in recent years (\cite{nW4HY8Nq}}), for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \cite{HqZ3r6J3}. RDF 1.1 formats use Unicode to support IRIs \cite{rqjuXkEf}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to). By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \cite{rnzJ4khD}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.

@@

Established Semantic Web services and standards include SPARQL (\cite{xww8fb1B}}) (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).

@@

Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies. Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS (\cite{15gQDya5B}}) (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties). Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \cite{FDDmgO6s} .

@@

With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources. This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \cite{9CdJnsoJ}. Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \cite{AGHR9oyr}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \cite{5SuwLNdE}. Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \cite{TUmA4KQT} , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS (\cite{ZwAcGQKY}}), OWL (\cite{1p4IWJpI}}), OBO \cite{1DalZuKct}, SKOS (\cite{15gQDya5B}})), hash vs slash, HTTP status codes and PID redirection strategies \cite{rnzJ4khD}.

@@

The \textbf{Linked Data} concept \cite{19s3yyxDn} was kickstarted as a set of best practices (\cite{OGXYtGZ8}}) to bring the Web aspect back into focus. Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers. This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.

Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data. In addition to several such embedded \emph{microformats} (Open Graph (\cite{oN9D5LkD}}), RDFa (\cite{FKIF6ApB}}), Microdata (\cite{14Utscfe4}})) we find JSON-LD (\cite{X5SWPqSU}}) as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications. JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites (\cite{j3SqQLBR}}).

Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\cite{BFMoe8FC}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL (\cite{GDrm9dgX}}) or ShEx (\cite{YUzogaV}})) can be used to validate RDF Data (\cite{ZeOAPNHa}; \cite{S5BI6Mwx}) before consuming it programmatically, or reshaping data to fit other models. While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts \cite{o596f2JM}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.

@@

In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed. These have shown interoperability. The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles (\cite{hxtgCGjY}}) to embrace openness, transparency, and broad consensus. In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference \cite{jrZe6Esu}. This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP (\cite{p7nwRj8E}}).

It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary. For example, DOIs are mostly resolved on the web (\cite{1H9iQhQYq}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail (\cite{68fzID2k}}).

@@

The FAIR Digital Object guidelines \cite{RwvirqWg} sets out recommendations for FDO implementations. In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \cite{13TcbsZF6} and using Linked Data Platform \cite{17OHlMRQA}, as proposed by (\cite{7szz7dwO}}).

@@

\begin{itemize}
\tightlist
\item
  G1 and G2 call for stability and trustworthiness. While the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
\item
  Machine actionability (G4, G6) is a core feature of both FDOs and Linked Data. Conceptually they differ in the which way types and operations are discovered, with FDO seemingly more rigorous. In practice, however, we see that DOIP also relies on dynamic discovery of operations and that operation expectations for types (FDOF7) have not yet been defined.
\item
  FDO proposes that types can have additional operations beyond CRUD (FDOF5, FDOF6), while Linked Data mainly achieves this with RESTful patterns using CRUD on additional resources, e.g.~\texttt{order/152/items}. These are mainly stylistics but affect the architectural view -- FDOs have more of an object-oriented approach.
\item
  FDO puts strong emphasis on the use of PIDs (FDOF1, FDOF2, FDOF3, FDOF5), but in current practice DOIP use local types, local extended operations (FDOF5) and attributes (FDOF4) that are not bound to any global namespace.
\item
  Linked Data have a strong emphasis on semantics (FDOF8), and metadata schemas developed by community agreements (FDOF10). FDO types need to be made reusable across servers.
\item
  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use. Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} (\cite{9T4j3N4e}})
\item
  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives, LDP has specific native support for containers.
\item
  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in table \ref{sec:fair-compare}).
\item
  DOIP supports multiple chunks of data for an object (FDOF3), while Linked Data can support content-negotiation. In either case it can be unclear to clients what is the meaning or equivalence of any additional chunks.
\end{itemize}

@@

We argue that the Web, with its now ubiquitous use of REST API \cite{174AwcFUL}, can be compared as a similar global middleware. Note that while early moves for developing Semantic Web Services \cite{WzPECPDP} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} (\cite{11dInoyVF}}).

@@

In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines \cite{RwvirqWg} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} (\cite{PR-RequirementSpec-2.0}}).

Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform (\cite{7szz7dwO}}) comply with the FAIR principles \cite{6DjakjNS}. Here we've used the RDA's FAIR Data Maturity Model \cite{1GrKTFaK2} as it has decomposed the FAIR principles to a structured list of FAIR indicators \cite{UzQhqk0M}, importantly considering \emph{Data} and \emph{Metadata} separately. In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.

@@

The FAIR Digital Object approach raises many important points for Linked Data practictioners.
At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake (\cite{kYjzRJ0H}}).
However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.

>>>>

The FAIR Digital Object Forum \cite{DMEBhHID} working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}. As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public. As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:

@@

\textbf{Upload of FDO} \cite{1H7cmmr69} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve). ResourceSync \cite{eS5y9TRh} is suggested as one type of service to list FDOs. This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).

@@

The concept of \textbf{FAIR Digital Objects} \cite{IHLT6hye} has been introduced as way to expose research data as active objects that conform to the FAIR principles \cite{6DjakjNS}. This builds on the \emph{Digital Object} (DO) concept \cite{11MnuwJ4l}, first introduced in 1995 \cite{3Uqe3fuK} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles. DO was the inspiration for the ITU X.1255 framework \cite{103Hw8H43} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 \cite{16uB3jxpa}.

@@

\begin{itemize}
\tightlist
\item
  Building on the Digital Object concept, using the simplified DOIP v2 specification \cite{13TcbsZF6}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0, see \cite{xuQKRRx5}} (usually TCP/IP over TLS). The main DOIP operations are retrieving, creating and updating digital objects. These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}. FDO types are registered in the local Cordra instance, where they are specified using JSON Schema \cite{15EZ2D0Rm} and PIDs are assigned using the Handle system. Several type registries have been established.
\item
  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
\item
  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
\end{itemize}

@@

Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) \cite{16WGmCih8} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format \cite{16wvFH4nh} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).

While triple-based knowledge representations were not new \cite{NxvhwSmG}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs \cite{RRtwPF7Y} are generalised forms of URLs that include locator-less identifiers
  such as ISBN book numbers (URNs). The distinction between locator-full and locator-less identifiers have weakened in recent years \cite{nW4HY8Nq}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \cite{HqZ3r6J3}. RDF 1.1 formats use Unicode to support IRIs \cite{rqjuXkEf}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to). By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \cite{rnzJ4khD}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.

@@

Established Semantic Web services and standards include SPARQL \cite{xww8fb1B} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).

@@

Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies. Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS \cite{15gQDya5B} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties). Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \cite{FDDmgO6s} .

@@

With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources. This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \cite{9CdJnsoJ}. Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \cite{AGHR9oyr}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \cite{5SuwLNdE}. Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \cite{TUmA4KQT} , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \cite{ZwAcGQKY}, OWL \cite{1p4IWJpI}, OBO \cite{1DalZuKct}, SKOS \cite{15gQDya5B}), hash vs slash, HTTP status codes and PID redirection strategies \cite{rnzJ4khD}.

@@

The \textbf{Linked Data} concept \cite{19s3yyxDn} was kickstarted as a set of best practices \cite{OGXYtGZ8} to bring the Web aspect back into focus. Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers. This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.

Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data. In addition to several such embedded \emph{microformats} (Open Graph \cite{oN9D5LkD}, RDFa \cite{FKIF6ApB}, Microdata \cite{14Utscfe4}) we find JSON-LD \cite{X5SWPqSU} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications. JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites \cite{j3SqQLBR}.

Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\cite{BFMoe8FC}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL \cite{GDrm9dgX} or ShEx \cite{YUzogaV}) can be used to validate RDF Data (\cite{ZeOAPNHa}; \cite{S5BI6Mwx}) before consuming it programmatically, or reshaping data to fit other models. While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts \cite{o596f2JM}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.

@@

In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed. These have shown interoperability. The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles \cite{hxtgCGjY} to embrace openness, transparency, and broad consensus. In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference \cite{jrZe6Esu}. This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP \cite{p7nwRj8E}.

It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary. For example, DOIs are mostly resolved on the web \cite{1H9iQhQYq} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \cite{68fzID2k}.

@@

The FAIR Digital Object guidelines \cite{RwvirqWg} sets out recommendations for FDO implementations. In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \cite{13TcbsZF6} and using Linked Data Platform \cite{17OHlMRQA}, as proposed by \cite{7szz7dwO}.

@@

\begin{itemize}
\tightlist
\item
  G1 and G2 call for stability and trustworthiness. While the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
\item
  Machine actionability (G4, G6) is a core feature of both FDOs and Linked Data. Conceptually they differ in the which way types and operations are discovered, with FDO seemingly more rigorous. In practice, however, we see that DOIP also relies on dynamic discovery of operations and that operation expectations for types (FDOF7) have not yet been defined.
\item
  FDO proposes that types can have additional operations beyond CRUD (FDOF5, FDOF6), while Linked Data mainly achieves this with RESTful patterns using CRUD on additional resources, e.g.~\texttt{order/152/items}. These are mainly stylistics but affect the architectural view -- FDOs have more of an object-oriented approach.
\item
  FDO puts strong emphasis on the use of PIDs (FDOF1, FDOF2, FDOF3, FDOF5), but in current practice DOIP use local types, local extended operations (FDOF5) and attributes (FDOF4) that are not bound to any global namespace.
\item
  Linked Data have a strong emphasis on semantics (FDOF8), and metadata schemas developed by community agreements (FDOF10). FDO types need to be made reusable across servers.
\item
  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use. Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} \cite{9T4j3N4e}
\item
  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives, LDP has specific native support for containers.
\item
  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in table \ref{sec:fair-compare}).
\item
  DOIP supports multiple chunks of data for an object (FDOF3), while Linked Data can support content-negotiation. In either case it can be unclear to clients what is the meaning or equivalence of any additional chunks.
\end{itemize}

@@

We argue that the Web, with its now ubiquitous use of REST API \cite{174AwcFUL}, can be compared as a similar global middleware. Note that while early moves for developing Semantic Web Services \cite{WzPECPDP} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} \cite{11dInoyVF}.

@@

In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines \cite{RwvirqWg} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} \cite{PR-RequirementSpec-2.0}.

Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform \cite{7szz7dwO} comply with the FAIR principles \cite{6DjakjNS}. Here we've used the RDA's FAIR Data Maturity Model \cite{1GrKTFaK2} as it has decomposed the FAIR principles to a structured list of FAIR indicators \cite{UzQhqk0M}, importantly considering \emph{Data} and \emph{Metadata} separately. In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.

@@

The FAIR Digital Object approach raises many important points for Linked Data practictioners.
At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \cite{kYjzRJ0H}.
However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
